### 2014-04-28 text/Chapter03.tex
Finish BLL, start working on structural dataflow (Chapter 3)

\caption{Structural coeffect liveness analysis}
\label{fig:applications-struct-bll}
\vspace{-1em}
\end{figure}

@@

\paragraph{Semantics.}
In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
$\lambda$-calculus language to encode the model. More formally, this could be, for example,
a Cartesian closed category. In that model, we can reuse variables arbitrarily and so it is not
a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
linear logic in an (ordinary) linear logic where variables can be used at most once.

@@

The rules that define the semantics of bounded calculus are mostly the same as (or easy to adapt 
from) the semantic rules of liveness in Figure~\ref{fig:applications-struct-livesem}. The ones that
differ are those that use sequential composition (application and let binding) and the contraction
rule, which represents pointwise composition. The bounded reuse calculus rules look as follows:

\begin{equation*}
\hspace{-0.5em}
\tag{\emph{app}}
\begin{array}{l}
\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
\end{array}
\end{equation*}

\begin{equation*}
\hspace{-0.5em}
\tag{\emph{contr}}
\begin{array}{l}
\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} =\\[-0.25em] 
\qquad  \lambda(\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}+\cclrd{t}}), \mathbf{v_2}) \rightarrow \\[-0.25em]
\qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
\qquad\qquad\qquad (\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{t}}), (x_{\cclrd{t}+1}, \ldots, x_{\cclrd{t}+\cclrd{s}} ), \mathbf{v_2})
\end{array}
\end{equation*}

\paragraph{Bounded linear logic.}

Bounded linear logic \cite{logic-bounded} restricts well-typed terms to polynomial-time algorithms.  
This is done by limiting the number of times a value (proposition) can be used. An assumption
$!_k A$ means that a variable can be used at most $k$ times. 
We attach annotations to the whole context rather than individual assumptions and so a context
$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.

Bounded linear logic includes explicit weakening and contraction rules
that affect the multiplicity. Following the original logical style (but with our notation), these are written as:
\[
\inference
  {\coctx{\Gamma}{\aclrd{R}} \vdash \tau}
  {\coctx{\Gamma,\sigma}{\aclrd{R} \cons \alift{0}} \vdash \tau} 
\quad
\inference
  {\coctx{\Gamma_1,\sigma,\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
  {\coctx{\Gamma_1,\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
\]
The context $\coctx{\Gamma}{\aclrd{R}}$ includes a \emph{coeffect annotation} $\aclrd{R}$ which is a vector
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$ of the same length as $\Gamma$ (a side-condition omitted for brevity).
In weakening (left), unused propositions are annotated with $0$ (no uses), while in contraction (right), multiple 
occurrences of a proposition are joined by adding the number of uses.

\paragraph{Summary.}

We really need just: $\ast$, $+$, $1$, $0$, $\leq$

composing flat and structural - meta vs. language semantics

@@

\subsection{Data-flow languages (revisited)}
When discussing data-flow languages in the previous section, we said that the context provides 
past values of variables. This can be viewed as a flat contextual property (the context needs
to keep all past values), but we can also view it as a structural property. Consider the following
example:
%
\begin{equation*}
\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
\end{equation*}
%
The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}.

@@

% % % % % %

\paragraph{Dataflow and data access}

Dataflow languages such as Lucid \cite{wadge1985lucid} describe computations over \emph{streams}.
An expression is re-evaluated when new inputs are available (push) or when more output
is demanded (pull).  In causal dataflow, programs can access past
values of a stream. We consider a language where $\kvd{prev}~e$
returns the previous value of $e$, where 
$\kvd{prev}~(\kvd{prev}~e)$ therefore returns the second past value.

An implementation of causal dataflow may cache past values of
variables as an optimisation. The question is, how many past values
should be cached?  This can be approximated by a coeffect system.

@@

\begin{figure}[t]
\[
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\ident{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{Q}} \vdash \subst{e}{y,z}{x} : \tau}
\]
\[
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} + \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
\]
\[
\tyrule{var}{}
  {\coctx{x\!:\!\tau}{\alift{0}} \vdash x : \tau}
\quad\;
\tyrule{prev}
  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
  {\coctx{\Gamma}{1 + \aclrd{R}} \vdash \kvd{prev}~e : \tau}  
\]
\caption{type and coeffect system for dataflow caching}
\label{fig:dataflow-coeff}
\end{figure}

>>>>

\caption{Structural coeffect bounded reuse analysis}
\label{fig:applications-struct-bll}
\vspace{-1em}
\end{figure}

@@

\paragraph{Semantics.}
In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian closed 
category. In that model, we can reuse variables arbitrarily and so it is not
a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
linear logic in an (ordinary) linear logic where variables can be used at most once.

@@

The rules that define the semantics of bounded calculus are mostly the same as (or easy to adapt 
from) the semantic rules of liveness in Figure~\ref{fig:applications-struct-livesem}. The ones that
differ are those that use sequential composition (application and let binding) and the contraction
rule, which represents pointwise composition.

In the following, we use variable names $\mathbf{v}_i$ for context containing multiple variables (where 
each variable may be available multiple times), \ie~have a type $\tau_1^{\cclrd{r_1}}\times\ldots\times\tau_m^{\cclrd{r_m}}$;
We do not explicitly write the sizes of these vectors (number of variables in a context; number of 
instances of a variable) as these are clear from the coeffect annotations. We assume that $\Gamma_2$ 
contains $n$ variables and that $\aclrd{s}=\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}$:
%
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{contr}}
\begin{array}{l}
\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} =\\[-0.25em] 
\qquad  \lambda(\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}+\cclrd{t}}), \mathbf{v_2}) \rightarrow \\[-0.25em]
\qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
\qquad\qquad\qquad (\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}}), (x_{\cclrd{s}+1}, \ldots, x_{\cclrd{s}+\cclrd{t}} ), \mathbf{v_2})
\end{array}
\end{equation*}
%
\begin{equation*}
\hspace{-0.5em}
\tag{\emph{app}}
\begin{array}{l}
\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = \\[-0.25em]
  \qquad\lambda (\mathbf{v_1}, ((x_{1,1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_1}), \ldots, (x_{n,1}, \ldots, x_{n,\cclrd{t}\ast\cclrd{s}_n}) ) \rightarrow \\[-0.25em]
  \qquad\qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
  \qquad\qquad\kvd{let}~\mathbf{y}_1 = ((x_{1,1}, \ldots, x_{1,\cclrd{s}_1}), \ldots, (x_{n,1}, \ldots, x_{1,\cclrd{s}_n})  )\\[-0.25em]
  \qquad\qquad\kvd{let}~\ldots \\[-0.25em]
  \qquad\qquad\kvd{let}~\mathbf{y}_{ \cclrd{t} } = ((x_{1,(\cclrd{t}-1)\ast\cclrd{s}_1 + 1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_1}), \ldots, \\[-0.25em]
  \hspace{8.2em}                                         (x_{n,(\cclrd{t}-1)\ast\cclrd{s}_n + 1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_n})  )\\[-0.25em]
  \qquad\qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}~\mathbf{y}_1, \ldots, 
    \sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}~\mathbf{y}_{\cclrd{t}})
\end{array}
\end{equation*}
%
In the (\emph{contr}) rule, the semantic function is called with $\cclrd{s}+\cclrd{t}$ copies of
a value for the $x$ variable. The values are split between $\cclrd{s}$ and $\cclrd{t}$ separate
copies of variables $y$ and $z$, respectively.

The (\emph{app}) rule is similar in that it needs to split the input variable context. However,
it needs to split values of multiple variables -- in $x_{i,j}$, the index $i$ stands for an
index of the variable while $j$ is an index of one of multiple copies of the value. In the
semantic function, the second
part of the context consists of $n$ variables where the multiplicity of each value is specified 
by the annotation $\cclrd{s}_i$ multiplied by $\cclrd{t}$. The rule needs to evaluate the argument 
$e_2$ $\cclrd{t}$-times and each call requires $\cclrd{s}_i$ copies of the $i^\textnormal{th}$ 
variable. To do this, we create contexts $\mathbf{y}_1$ to $\mathbf{y}_{\cclrd{t}}$, each containing
$\cclrd{s}_i$ copies of the variable (and so we require $\cclrd{s}_i\ast\cclrd{t}$ copies of each 
variable). Note that the contexts are created such that each value is used exactly once.

It is worth noting that the (\emph{var}) rule requires exactly one copy of a variable and so 
the system tracks precisely the number of uses. However, the (\emph{sub}) rule lets us 
ignore additional copies of a value. Thus, permitting (\emph{sub}) rule is only possible if the
underlying model is \emph{affine} rather than \emph{linear}.

\paragraph{Bounded linear logic.}
The system presented in this section differs from bounded linear logic (BLL) \cite{logic-bounded}.
Using the terminology from Section~\ref{sec:path-sem-contextdep}, our system is written in the
\emph{language semantics} style, while BLL is written in the \emph{meta-language} style.

This means that the terms and types of our system are the terms and types of ordinary 
$\lambda$-calculus, with the only difference that functions carry coeffect annotations.
In BLL, the language of types is extended with a type constructor $!_k A$ (where $A$ is
a proposition, corresponding to a type $\tau$ in our system). The type denotes a value
$A$ that can be used at most $k$ times.

As a result, BLL does not need to attach additional annotation to the variable context 
as a whole. The requirements are attached to individual variables and so our context
$\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$ corresponds
to a BLL assumption $!_{k_1} A_1, ..., !_{k_n} A_n$.

Using the formulation of bounded logic (and omitting the terms), the weakening and
contraction rules are written as follows:
%
\[
\tyrule{weak}
  {\Gamma \vdash B}
  {\Gamma, !_0 A \vdash B}
\quad
\tyrule{contr}
  {\Gamma, !_n A, !_m A \vdash B}
  {\Gamma, !_{n+m} A \vdash B}
\]
%
The system captures the same idea as the structural coeffect system presented above.
Variable access in bounded linear logic is simply an operation that produces a value
$!_n A$ and so the system further introduces \emph{dereliction} rule which lets us 
treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
semantics} and \emph{meta-language} and also revisit the BLL example in 
Chapter~\ref{ch:meta}.

\paragraph{Summary.}
Comparing the structural coeffect calculus for tracking liveness and for bounded variable reuse
reveals which parts of the systems differ and which parts are shared. In particular, both systems
use the same vector operations ($\times$, $\alift{\cclrd{\textnormal{--}}}$) and also share the
lambda abstraction rule (\emph{abs}). They differ in the primitive values used to annotate used
and unused variables (\cclrd{L}, \cclrd{D} and $1$, $0$, respectively) and in the operators used
for sequential composition and contraction ($\sqcup$, $\sqcap$ and $\ast$, $+$, respectively).
The algebraic structure capturing these operators is developed in Chapter~\ref{ch:structural}.

The brief overview of bounded linear logic shows an alternative approach to tracking properties
related to individual variables -- we could attach annotations to the variables themselves 
rather than attaching a \emph{vector} of annotations to the entire context. Our approach has
two benefits -- it lets you unify flat and structural systems (Chapter~\ref{ch:unified}) and 
it also makes it possible to build composed systems that mix both flat and structural properties.

@@

\subsection{Data-flow languages revisited}

When discussing data-flow languages in the previous section, we said that the context provides 
past values of variables. This can be viewed as a flat contextual property (the context needs
to keep all past values), but we can also view it as a structural property. Consider the following
example:
%
\begin{equation*}
\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
\end{equation*}
%
The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}.

@@

@@

% --------------------------------------------------------------------------------------------------
% variable annotated with 0 (not 1)
% sub and abs omitted
% app uses +, not \ast
% let is default

\begin{figure}[t]
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{0}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{prev}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{1 + \aclrd{\textbf{r}}} \vdash \kvd{prev}~e : \tau}  
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} + \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{0} }} \vdash e : \sigma} 
\end{equation*}
\begin{equation*}
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\textnormal{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
\end{equation*}

\caption{Structural coeffect bounded reuse analysis}
\label{fig:applications-struct-df}
\vspace{-1em}
\end{figure}
